An Efficient Feedback Calibration Algorithm for Direct Imaging Radio Telescopes

### 2016-07-11 cal_demo/ms.tex
Finished JDBs comments

% Abstract of the paper
\begin{abstract}
We present the E-field Parallel Imaging Calibration (EPICal) algorithm, which addresses the 
need for a real-time calibration method for direct imaging radio astronomy correlators. Direct 
imaging involves a spatial fast Fourier transform of antenna \textbf{signals}, alleviating 
\textbf{an 
$\mathcal{O}(\Nant^2)$ computational bottleneck typical in radio correlators, and yielding a more gentle $\mathcal{O}(\Nant \log_2 
\Nant)$ scaling.} \textbf{This} can save orders of magnitude in computation cost for next generation arrays 
consisting of hundreds to thousands of antennas. However, because signals are mixed in the 
correlator, gain correction must be applied \textbf{prior to correlation, rather than on visibilities post-correlation.} We develop the EPICal algorithm 
to form gain solutions in real time without ever forming visibilities. This method scales as the 
number of antennas, and produces results comparable to those from visibilities. Through 
simulations and application to Long Wavelength Array data we show this algorithm is a 
promising solution for next generation instruments.
\end{abstract}

@@

A challenge common to all direct imaging algorithms is calibration of the antenna gains. With a 
traditional \textbf{cross correlator}, pair-wise visibilities are written to disk and used to calibrate 
\textbf{after observing and before further processing} such as imaging. However, a direct imaging correlator mixes 
the signals from all antennas before averaging and writing to disk, making calibration a 
requirement at the front end, before imaging and averaging. Previous solutions have involved 
applying calibration solutions generated from a parallel FX correlator \citep{zhe14, fos14}, or 
integrating a dedicated FX correlator which periodically formed the full visibility matrix to solve 
for gains \citep{wij09,dev09}. While these solutions were sufficient to enable the exploration of 
FFT correlators and beamformers, they will not scale to future arrays with $\Nant \gtrsim 10^3$.

@@

\textbf{The physical parameters of our simulation can be arbitrarily scaled to other cases. However, 
we provide typical units for an MWA observation as an example. Our simulated antenna 
array consists of} the inner 51 antennas of the MWA layout \citep{bea12}, within a bounding 
box of 150~m. The \textbf{antenna aperture illumination pattern} used is a 4.4~m square tophat; \textbf{a very rough 
approximation to the square shaped tiles of the MWA.} Because our algorithm treats frequency 
channels independently, we simulate only one channel. For context we treat this as \textbf{a channel
at 150~MHz with 40 kHz bandwidth, and adopt a sampling period of 25~$\mu$s. The simulated signal 
consists of 10 random point sources with flux densitities 0.5~Jy~$\lesssim S \lesssim$~1~Jy 
within the main lobe of the primary beam.}

For our unknown gains, we create a set of random complex numbers where the amplitude is 
drawn from an approximately gaussian distribution centered around \textbf{unity} with width 0.25, and 
completely random phase. These are our ``true gains", and we apply them to the frequency
domain simulated antenna electric field \textbf{measurements} as in equation~\ref{eq:apply_gain}. Our analysis is 
blind to these values until the end of the process to check accuracy. The gain estimates are 
initialized with unity, $g^{(0)}_a=1$.

We next process and image \caliter~time \textbf{samples} (10~ms). We also form the correlations, 
\Cna[0], used in our calibration loop. The pixel used for the correlation is the source with the 
largest apparent flux (intrinsic flux attenuated by the primary beam). These correlation values 
are used to update the gain estimates. \textbf{For this simple demonstration we create perfect model 
visibilities of our 10 simulated sources to be used in equation~\ref{eq:damping}. 
The updated gain estimates are} used to calibrate the following 
\caliter~time samples. Through experimentation we found a damping factor of $\damp=0.35$ 
resulted in the quickest convergence in this simulation.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig4.pdf}
\caption{Gain estimate amplitudes as a function of iteration for simulated calibration. Again, 
each line represents an antenna in the 51 MWA antenna sample. The gain estimates were 
initialized randomly, while the true values were unity. After about \textbf{12} iterations we see the 
calibration loop has settled around the correct values, with only noise remaining.}
\label{fig:sim_amp}
\end{center}
\end{figure}

@@

The results of our simulations are shown in Fig.~\ref{fig:errors}. The gain errors from our 
EPICal method are shown with solid lines, while the dashed lines represent the visibility-based 
calibration. The thick lines represent the results of using a full sky model, and the thin lines are 
from the single point source model.  In the case of the full sky model, we see the errors trend 
downward with longer integration times, as expected. The EPICal errors are slightly higher than 
the visibility-based errors, on average about 23\% difference. This is not surprising as EPICal 
only uses the information in a single pixel, while the visibilities use all information from the full 
sky model. However, we see that with this perfect model the errors do behave like noise
\textbf{in the sense that they integrate down with longer integrations with the expected $t^{-1/2}$ 
trend. }
The same level of \textbf{gain error as visibility based solutions} can be achieved with a longer integration time (or larger damping 
factor).

@@

When using the single point source model, all calibration methods in Fig.~\ref{fig:errors} trend 
downward until they reach an error floor due to the confusion sources that were not modeled. 
The EPICal and visibility-based solutions reach a similar floor, but EPICal achieves a 
marginally lower level (on average 5\% lower). This can be attributed to the same reason 
EPICal underperformed with the full sky model: EPICal only used a single pixel in the sky to 
form its solutions. When using the full sky model, it was \textbf{under utilizing} the information in the 
rest of the sky. But when the sky model only contains the bright point source, the single pixel 
used in EPICal is the location where this model is most accurate, effectively down weighting 
the pixels with incomplete sky model. In the regime typical of HERA discussed above, EPICal's 
error floor was 10\% lower than that of visibility-based solutions.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig8.pdf}
\caption{Gain amplitudes as a function of calibration iteration for an LWA TBN observation. The 
gains estimates were initialized with amplitude 0.25 after inspection of the electric field values 
compared to our model sky. The solutions are noisy, but flat. The range in amplitudes is due to 
the non-uniformity of cables between the LWA antennas and receivers.
}
\label{fig:data_amp}
\end{center}
\end{figure}

Figure~\ref{fig:data_images} shows the improvement in the images due to our calibration. The 
left panel shows the uncalibrated image integrated over 51.2 ms, \textbf{83 kHz.} Cyg~A is prominent 
near the center of the image, and Cas~A is also clearly visible in the upper right. The middle 
panel shows the image produced after calibration with identical integration time and bandwidth. 
The sidelobes throughout the image are significantly suppressed, and the galactic plane is 
much more evident, despite only modeling Cyg~A and Cas~A. We also note that the feature 
just to the right of Cyg~A is dimmer in the calibrated image, better matching the expected flux 
from the GSM. 
\textbf{For reference we show the GSM in the right panel of 
Fig.~\ref{fig:data_images}, convolved by the LWA point spread function and weighted
by  two factors of the primary beam to match the holographic frame of the data images.}

\begin{figure*}
\begin{center}
\includegraphics[width=0.3\linewidth]{fig9a.pdf}
\includegraphics[width=0.3\linewidth]{fig9b.pdf}
\includegraphics[width=0.3\linewidth]{fig9c.pdf}
\caption{Images produced before (left) and after (middle) calibrating LWA data. These images 
were produced with \textbf{83 kHz} bandwidth and 51.2 ms integration. The calibrated image shows 
significant reduction in \textbf{rumble} throughout, while retaining the prominent Cyg~A and 
Cas~A sources. The galactic plane is also substantially more evident after calibration. For 
reference, the GSM is shown \textbf{in the 
right panel convolved with the LWA point spread function and weighted by two factors
of the primary beam to place it in the same holographic frame as the data images.} While much of the GSM is visible in the calibrated image, we 
note that the model used to calibrate was only the two bright sources, Cyg~A and Cas~A.
}
\label{fig:data_images}
\end{center}
\end{figure*}

To compare the image qualities quantitatively we compute the dynamic range, defined as the 
peak of the image over the noise level. We estimate the noise level as the median of the 
absolute deviation of the image. With this metric we find the \textbf{uncalibrated 
and} calibrated images to have \textbf{dynamic ranges of 88.5 and 137, respectively, a 55\% 
improvement.}

@@

\textbf{Improving sky model.} In this work we used a perfect sky model (for simulation), or a 
very simple sky model (for LWA data). In principle the images produced by the EPIC correlator 
can be used to improve the sky model used in calibration -- similar to a major loop in 
self-calibration. This can be especially useful for compact, widefield arrays which require a model of 
both compact and diffuse sources over a large patch of sky. In addition to requiring the sky 
model, the calibration also requires this model to be attenuated by the primary beam, which 
can be difficult to measure at the precision necessary \citep[e.g.][]{neb15,vir14,thy15b}. 
However, the direct imaging correlator provides exactly the \textbf{product of the sky and beam} necessary in real time, and 
can be iterated over to improve the images and gain solutions, \textbf{leaving final beam
correction to post-processing as in traditional processing.}

>>>>

% Abstract of the paper
\begin{abstract}
We present the E-field Parallel Imaging Calibration (EPICal) algorithm, which addresses the 
need for a real-time calibration method for direct imaging radio astronomy correlators. Direct 
imaging involves a spatial fast Fourier transform of antenna \textbf{signals}, alleviating 
\textbf{an 
$\mathcal{O}(\Nant^2)$ computational bottleneck typical in radio correlators, and yielding a more gentle $\mathcal{O}(\Nant \log_2 
\Nant)$ scaling.} \textbf{This} can save orders of magnitude in computation cost for next generation arrays 
consisting of hundreds to thousands of antennas. However, because signals are mixed in the 
correlator, gain correction must be applied \textbf{prior to correlation, rather than on visibilities post-correlation.} We develop the EPICal algorithm 
to form gain solutions in real time without ever forming visibilities. This method scales as the 
number of antennas, and produces results comparable to those from visibilities. 
\textbf{We use simulations to demonstrate the EPICal technique and study the noise properties of
our gain solutions, showing they are similar to visibility based solutions in realistic 
situations.
By applying EPICal to Long Wavelength Array data we achieve a 65\% dynamic range
improvement compared to uncalibrated images, showing this algorithm is a promising
solution for next generation instruments.}
\end{abstract}

@@

A challenge common to all direct imaging algorithms is calibration of the antenna gains. With a 
traditional \textbf{cross correlator}, visibilities are written to disk and used to calibrate 
\textbf{after observing and before further processing} such as imaging. However, a direct imaging correlator mixes 
the signals from all antennas before averaging and writing to disk, making calibration a 
requirement at the front end, before imaging and averaging. Previous solutions have involved 
applying calibration solutions generated from a parallel FX correlator \citep{zhe14, fos14}, or 
integrating a dedicated FX correlator which periodically formed the full visibility matrix to solve 
for gains \citep{wij09,dev09}. While these solutions were sufficient to enable the exploration of 
FFT correlators and beamformers, they will not scale to future arrays with $\Nant \gtrsim 10^3$.

@@

\textbf{Our simulated antenna 
array consists of} the inner 51 antennas of the MWA layout \citep{bea12}, within a bounding 
box of 150~m. 
\textbf{The physical parameters of our simulation can be arbitrarily scaled to other cases. However, 
we provide typical units for an MWA observation as an example. 
The antenna aperture illumination pattern} used is a 4.4~m square tophat; \textbf{a very rough 
approximation to the square shaped tiles of the MWA.} Because our algorithm treats frequency 
channels independently, we simulate only one channel. For context we treat this as \textbf{a channel
at 150~MHz with 40 kHz bandwidth, and adopt a sampling period of 25~$\mu$s. The simulated signal 
consists of 10 random point sources with flux densitities 0.5~Jy~$\lesssim S \lesssim$~1~Jy 
within the main lobe of the primary beam.}

For our unknown gains, we create a set of random complex numbers where the amplitude is 
drawn from \textbf{a gaussian distribution centered around unity with width 0.25, and truncated at zero. The gain phases are drawn from a uniform distribution between $\pm \pi$. }
These are our ``true gains", and we apply them to the frequency
domain simulated antenna electric field \textbf{measurements} as in equation~\ref{eq:apply_gain}. Our analysis is 
blind to these values until the end of the process to check accuracy. The gain estimates are 
initialized with unity, $g^{(0)}_a=1$.

We next process and image \caliter~time \textbf{samples} (10~ms). We also form the correlations, 
\Cna[0], used in our calibration loop. The pixel used for the correlation is the source with the 
largest apparent flux (intrinsic flux attenuated by the primary beam). These correlation values 
are used to update the gain estimates. \textbf{We create perfect model 
visibilities of our 10 simulated sources to be used in equation~\ref{eq:damping}. 
The updated gain estimates are} used to calibrate the following 
\caliter~time samples. Through experimentation we found a damping factor of $\damp=0.35$ 
resulted in the quickest convergence in this simulation.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig4.pdf}
\caption{\textbf{Same as Fig.~\ref{fig:sim_phase}, but for gain amplitudes.
Each} line represents an antenna in the 51 MWA antenna sample. 
\textbf{The relative amplitude plotted is the magnitude of the estimated gains divided by the true gains.}
After about \textbf{12} iterations we see the 
calibration loop has settled around the correct values, with only noise remaining.}
\label{fig:sim_amp}
\end{center}
\end{figure}

@@

The results of our simulations \textbf{for all four calibration methods are shown in Fig.~\ref{fig:errors}. }
In the case of the full sky model, we see the errors trend 
downward with longer integration times, as expected. The EPICal errors are slightly higher than 
the visibility-based errors, on average about 23\% difference. This is not surprising as EPICal 
only uses the information in a single pixel, while the visibilities use all information from the full 
sky model. However, we see that with this perfect model the errors do behave like noise
\textbf{in the sense that they integrate down with longer integrations with the expected $t^{-1/2}$ 
trend. }
\textbf{With longer integration time (or larger damping factor), EPICal can achieve the same 
level of gain error as visibility based solutions.}

@@

\textbf{In a more realistic situation, the observer will not have a perfectly complete model of the sky.
A common method is to instead model the brightest sources in the field,
or even a single dominant source.}
When using the single point source model, all calibration methods in Fig.~\ref{fig:errors} trend 
downward until they reach an error floor due to the confusion sources that were not modeled. 
The EPICal and visibility-based solutions reach a similar floor, but EPICal achieves a 
marginally lower level (on average 5\% lower). This can be attributed to the same reason 
EPICal underperformed with the full sky model: EPICal only used a single pixel in the sky to 
form its solutions. When using the full sky model, it was \textbf{under utilizing} the information in the 
rest of the sky. But when the sky model only contains the bright point source, the single pixel 
used in EPICal is the location where this model is most accurate, effectively down weighting 
the pixels with incomplete sky model. 
\textbf{In contrast, the visibility-based calibration method weights the apparent sky equally,
and experiences a more harsh penalty for missing sources in the model.}
In the regime typical of HERA discussed above, EPICal's 
error floor was 10\% lower than that of visibility-based solutions.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig8.pdf}
\caption{\textbf{Same as Fig.~\ref{fig:data_phase}, but for gain amplitudes.}
The 
gains estimates were initialized with amplitude 0.25 after inspection of the electric field values 
compared to our model sky. The solutions are noisy, but flat. The range in amplitudes is due to 
the non-uniformity of cables between the LWA antennas and receivers.
}
\label{fig:data_amp}
\end{center}
\end{figure}

Figure~\ref{fig:data_images} shows the improvement in the images due to our calibration. The 
left panel shows the uncalibrated image integrated over 51.2 ms, \textbf{58.6 kHz.} Cyg~A is prominent 
near the center of the image, and Cas~A is also clearly visible in the upper right. The middle 
panel shows the image produced after calibration with identical integration time and bandwidth. 
The sidelobes throughout the image are significantly suppressed, and the galactic plane is 
much more evident, despite only modeling Cyg~A and Cas~A. We also note that the feature 
just to the right of Cyg~A is dimmer in the calibrated image, better matching the expected flux 
from the GSM. 
\textbf{For reference we show the GSM in the right panel of 
Fig.~\ref{fig:data_images}, convolved by the LWA point spread function and weighted
by  two factors of the primary beam to match the holographic frame of the data images.}

\begin{figure*}
\begin{center}
\includegraphics[width=0.3\linewidth]{fig9a.pdf}
\includegraphics[width=0.3\linewidth]{fig9b.pdf}
\includegraphics[width=0.3\linewidth]{fig9c.pdf}
\caption{Images produced before (left) and after (middle) calibrating LWA data. These images 
were produced with \textbf{58.6 kHz} bandwidth and 51.2 ms integration. The calibrated image shows 
significant reduction in \textbf{rumble} throughout, while retaining the prominent Cyg~A and 
Cas~A sources. The galactic plane is also substantially more evident after calibration. For 
reference, the GSM is shown \textbf{in the 
right panel convolved with the LWA point spread function and weighted by two factors
of the primary beam to place it in the same holographic frame as the data images.} While much of the GSM is visible in the calibrated image, we 
note that the model used to calibrate was only the two bright sources, Cyg~A and Cas~A.
}
\label{fig:data_images}
\end{center}
\end{figure*}

To compare the image qualities quantitatively we compute the dynamic range, defined as the 
peak of the image over the noise level. We estimate the noise level as the median of the 
absolute deviation of the image. With this metric we find the \textbf{uncalibrated 
and} calibrated images to have \textbf{dynamic ranges of 80.4 and 133.1, respectively, a 65\% 
improvement.}

@@

\textbf{Improving sky model.} In this work we used a perfect sky model (for simulation), or a 
very simple sky model (for LWA data). In principle the images produced by the EPIC correlator 
can be used to improve the sky model used in calibration -- similar to a major loop in 
self-calibration. This can be especially useful for compact, widefield arrays which require a model of 
both compact and diffuse sources over a large patch of sky. 
\textbf{As with any sky-based calibration scheme, EPICal requires that the sky model be attenuated
by the primary beam,} which 
can be difficult to measure at the precision necessary \citep[e.g.][]{neb15,vir14,thy15b}. 
However, the direct imaging correlator provides exactly the \textbf{product of the sky and beam} necessary in real time, and 
can be iterated over to improve the images and gain solutions, \textbf{leaving final beam
correction to post-processing as in traditional processing.}
